Viscous Erosion in Porous Media

### 2018-03-03 docs/methods/erosion_methods.tex
NM

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arc length, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be handled in a 
numerically stable fashion, e.g.~via an exponential integrator or an implicit method.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary evolution in the {\thL} framework and regularization} 
\label{sec:thetaL}

To evolve the boundaries, we use the {\thL} framework, which offers certain advantages in stabilizing fluid-structure interaction problems \cite{hou-low-she1994}.

For numerical stability, we modify the erosion law \eqref{eqn:erosionModel} to include a term that depends on the local curvature $\kappa$. The idea is to prevent regions of very high
curvature from developing, as these regions would compromise the accuracy and stability of our fluid solver. We thus modify the normal velocity of each interface as follows
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1
  \right).
\end{align}
The second term on the right is the curvature-dependent modification, the strength of which is controlled by the parameter $\eps \ll 1$. Here, $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a boundary mean. This term causes regions of high curvature to recede more rapidly, thus smoothing any high-frequency oscillations that may exist. Since a closed body has mean curvature $\mean{\kappa} = 2\pi / L$, this smoothing term has mean zero and and thus preserves area. Therefore, the only source of material loss is the shear-stress itself. In addition, we have scaled the smoothing term on the mean absolute shear, $\mean{\atau}$, so that the ratio of smoothing-to-erosion always remains fixed, even as bodies vanish.

We now introduce the tangent angle, $\theta$, defined through the relation
\begin{align*}
\left( \pderiv{x}{s}, \pderiv{y}{s} \right) = \left(\cos \theta, \sin
\theta \right),
\end{align*}
where $s$ is arc length. We also introduce the normalized arc length $\alpha = 2 \pi s / L$. In terms of these new variables, the curvature can be expressed as
\begin{align*}
\kappa = \frac{2 \pi}{L} \pderiv{\theta}{\alpha}.
\end{align*}
Substitution into~\eqref{eqn:Vn} gives
\begin{align}
\label{eqn:Vn2}
\Vn = \atau +  \eps \mean{\atau}   \left(\pderiv{\theta}{\alpha} - 1
\right).
%\Vn = \atau + \frac{2 \pi \eps}{L \log \left(2\pi/L \right)}  \left(\pderiv{\theta}{\alpha} - 1 \right)
\end{align}

The main idea is to introduce an artificial tangential velocity that
keeps collocation points equally spaced with respect to arc length;
i.e.~the grid of $\alpha$ should always remain equally spaced. The
tangential velocity that does the trick satisfies~\cite{hou-low-she1994}
\begin{align*}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn}.
\end{align*}
The interface evolution is then given by
\begin{align*}
\dot{\xx}(t) = \Vn \nn + \Vs \ss.
\end{align*}
This evolution equation can be recast in the variables $\theta$ and $L$
as
\begin{align*}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\pderiv{\theta}{t} &= \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} +
\thalpha \Vs \right).
\end{align*}
Inserting~\eqref{eqn:Vn2} for the normal velocity gives
\begin{align}
\label{eqn:Levo1}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\label{eqn:thetaevo1}
\pderiv{\theta}{t} &= \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} +
\thalpha \Vs \right).
\end{align}
System \eqref{eqn:Levo1}--\eqref{eqn:thetaevo1} comprise the main evolution equations that we will solve numerically in Section~\ref{sec:timeStepping}.

In addition, it is necessary to track a reference point for each body, so that the variables $\theta$ and $L$ can be converted back into a cartesian grid. We choose to track the coordinates of the surface mean $\mean{\xx}$, which moves according to
\begin{align}
\label{eq:xsurf}
\frac{d}{dt} \mean{\xx} = \mean{ V_s \ss + V_n \nn }
\end{align}
Tracking the surface mean is somewhat simpler than tracking the center of mass, but either choice would be valid.

>>>>

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly (???) ---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arc length, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be handled in a 
numerically stable fashion, e.g.~via an exponential integrator or an implicit method.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary evolution in the {\thL} framework} 
\label{sec:thetaL}

To evolve the boundaries of eroding bodies, we will use the {\thL} framework which offers certain advantages in numerically stabilizing moving-interface problems \cite{hou-low-she1994}. Here, $\theta$ is the local tangent angle and $L$ the total perimeter of a given boundary. For a boundary parametrized by arc length $(x(s),y(s))$, the local tangent angle is defined through the relation $\left( \pdi{x}{s}, \pdi{y}{s} \right) = \left(\cos \theta, \sin
\theta \right)$.

For numerical stability, we modify the erosion law \eqref{eqn:erosionModel} to include a term that penalizes curvature $\kappa$. The idea is to prevent regions of extremely high curvature from developing, as these regions would compromise the accuracy and stability of our fluid solver. We thus modify \eqref{eqn:erosionModel} follows
\begin{align}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right).
\end{align}
The parameter $\eps \ll 1$ controls the strength of the curvature-penalization. This term causes regions of high curvature to recede more rapidly and thus smooths out any high-frequency oscillations that may exist. Above, $\mean{\cdot}$ indicates a boundary mean. Since a closed body has mean curvature $\mean{\kappa} = 2\pi / L$, the penalty term has mean zero and and thus preserves area. The only source of material loss is therefore the shear-stress itself. In addition, we have scaled the penalty term on the mean absolute shear, $\mean{\atau}$, so that the ratio of smoothing-to-erosion always remains fixed, even as bodies vanish.

In the following, we parameterize each boundary by normalized arc length, $\alpha = 2 \pi s / L$. In terms of $\alpha$, the curvature is given by
\begin{align}
\kappa = \frac{2 \pi}{L} \pderiv{\theta}{\alpha} \, ,
\end{align}
and substitution into~\eqref{eqn:Vn} gives
\begin{align}
\label{eqn:Vn2}
\Vn = \atau +  \eps \mean{\atau}   \left(\pderiv{\theta}{\alpha} - 1 \right) \, .
\end{align}

The boundary of each body will be represented by a grid of collocation points. If we were to simply evolve the interfaces under \eqref{eqn:Vn2}, this grid would become distorted, which, if unchecked, could lead to severe tangling and numerical instability. We therefore introduce an artificial tangential velocity $\Vs$ that always keeps the grid points equally spaced in arclength \cite{hou-low-she1994}. The appropriate tangential velocity satisfies 
\begin{align}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn} \, ,
\end{align}
The evolution of the interface is then given by the combination of the normal and tangential velocity
\begin{align}
\dot{\xx}(t) = \Vn \nn + \Vs \ss \, .
\end{align}
Since $\Vs$ is tangential, it does not modify the erosion law \eqref{eqn:Vn}, only the collocation grid.

When recast in the variables $\theta$ and $L$, the interface evolution becomes
\begin{align}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\pderiv{\theta}{t} &= \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} +
\thalpha \Vs \right).
\end{align}
Inserting~\eqref{eqn:Vn2} for the normal velocity gives
\begin{align}
\label{eqn:Levo1}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\label{eqn:thetaevo1}
\pderiv{\theta}{t} &= \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} +
\thalpha \Vs \right).
\end{align}
System \eqref{eqn:Levo1}--\eqref{eqn:thetaevo1} comprise the main evolution equations to be solved numerically in Section~\ref{sec:timeStepping}.

In addition, it is necessary to track a reference point for each body, so that the variables $\theta$ and $L$ can be converted back to cartesian coordinates. We track the coordinates of the surface mean $\mean{\xx}$, which moves according to
\begin{align}
\label{eq:xsurf}
\frac{d}{dt} \mean{\xx} = \mean{ V_s \ss + V_n \nn }
\end{align}
We note that tracking the center of mass is also a valid choice, but the surface mean gives a somewhat simpler formula.
